Gaia's constraints on the very massive (initially $\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-07-05 VFTS682_APJL.tex
updated sec.3 and caption/legend of fig.1

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to reside in the innermost cores. In this picture, field stars are primarily the result of the dissolution of dense groups.  However, a significant population of massive stars exists in relative isolation, far from dense clusters or OB associations and their origin remains matter of debate \citep{gvaramadze:12, lamb:16,ward:18}.   One hypothesis to explain the population of relatively isolated massive stars is that they formed in the field. The alternative hypothesis is that these massive stars were ejected from the clusters in which they formed.  Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the \SdM{ companion  star \citep[e.g.,][]{zwicky:57, blaauw:61, eldridge:11, renzo:18}}.

@@

% Observations with the Chandra X-Ray Observatory do not reveal an X-ray point source at the position of VFTS682 \cite{townsley:06}. Such a detection would be expected if the star hosted an accreting compact object or if the strong winds of VFTS682 are colliding with those of a stellar companion. This star is also relatively isolated in \emph{Spitzer} images \citep{meixner:06}, except for the extended source S10 interpreted as star formation triggered by the wind of VFTS682 itself by \cite{walborn:13}. 
% >>>>>>> df97f49fb6e094f5c7a3bd15402c437202858eaa

@@

\SdM{VFTS682 also appears to be relatively isolated in (near) infrared.   There nearest bright (near) infrared sources detected by \emph{Spitzer}  \citep{meixner:06} and resolved in the \emph{VISTA} Magellanic Clouds Survey \citep{cioni:11} are located at a distance of about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate that these nearby young stars may represent a case of starformation triggered by the the wind of VFTS682, but we chance alignment cannot be excluded.  }

% interpreted as star formation triggered by the wind of  VFTS682 itself by \cite{walborn:13}.   - SdM: too strong lets be more careful.  This is a controversial topic.  Walborn formulates more careful than you summarize him.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.

@@

%\SdM{\citet{bestenlehner:11} further report a wind mass loss rate of $10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, without correcting for the possible effect of clumping.  The helium surface  abundance is  $Y=0.45\, (0.49)$ \cite{bestenlehner:11} (\citealt{rubio-diez:17}).  IN TABLE ONLY TO SAFE SPACE}

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.}
  \label{fig:dist}
\end{figure*}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e., groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable. % We now include the correlation coefficient in the error
          % calculation, so no need for this paragraph? Gaia errors are correlated \citep[][]{lindengren:18}, but for this star the correlation coefficient is very small, $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$  = 0.026, where zero is for completely un-correlated data and one is for fully correlated data. We can thus safely treat the proper motion components as uncorrelated.

@@

\subsection{HST astrometry for VFTS682}
% observations; the first epoch (GO-12499; P.I.: D.~J.~Lennon)
% in October 3-8 2011 and the second epoch (GO-13359) in
% October 6-11 2014, both using the F775W filter.
% The second epoch was designed to match the pointing and
% orientation of the first epoch as exactly as possible so
% that the displacement measurements could be purely differential.

%For large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even exceed \emph{Gaia} DR2 data in quality. \SdM{Mathieu: do you have a ref for this?}

@@

%The errorbars are comparable with those we obtain from \emph{Gaia}.  However the direction and angle are different.

%ID     /HST relative mu_x/HST relative mu_y/HST error_mu_x/HST error_my_y/  HST F775W mag/Gaia B-mag?/Gaia relative mu_x/  Gaia relative mu_y/Gaia error_pmRA/Gaia error_pmDec

%682  -0.02   0.19 0.10 0.09 15.175 15.656   0.133   0.166 0.070   0.080

%The catalog from \citet{platais:18} lists the proper motion components of VFTS682 relative to the field, providing a
%measure of the motion of this star completely independent from
%\emph{Gaia}. They found proper
%motion components relative to the field
%$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
%  yr^{-1}}$ and
%$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
%  yr^{-1}}$, which added in quadrature give a relative proper motion of
%$\delta \mu^{HST} =0.20 \pm 0.15\masyr$.
% \begin{equation}
%   \label{eq:pm_around_HST}
%   \delta \mu^{HST} = \sqrt{\left(\delta\mu_\mathrm{RA}^{HST}\right)^2+\left(\delta\mu_\mathrm{DEC}^{HST}\right)^2}
%   = 0.20 \pm 0.15\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}

%These proper motion components can also be converted into projected
%velocities assuming a distance of 50\,kpc, obtaining $\delta
%v_\mathrm{RA}^{HST}=2\pm31\kms$ and $\delta
%v_\mathrm{RA}^{HST}=47\pm24\kms$% , corresponding to a two-dimensional
%% projected velocity of $47\pm24\kms$
%.
%Adding them in quadrature to the radial velocity
%measured by \citet{bestenlehner:11} we obtain a three-dimensional velocity of
%$v_\mathrm{pec}^{HST}=56 \pm 23  \kms$.
% \begin{equation}
%   \label{eq:speed_around_HST}
%   v_\mathrm{pec}^{HST} = \sqrt{\left(\delta v_\mathrm{RA}^{HST}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{HST}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 56 \pm 23
%   \kms \ .
% \end{equation}

@@

% At first glance the magnitude of the relative proper motions derived by Gaia,  $32\pm 21\kms$, and by HST, $47 \pm 24\kms$, appear consistent.  In principle they are consistent with the velocity

Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light shaded
blue). We highlight VFTS682 together with the two O-type
runaways \mbox{(re-)identified} by \cite{lennon:18} (VFTS16 and VFTS72,
respectively). We emphasize that selecting stars with small error bars
implicitly biases the sample toward fast moving objects.

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other and indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

The right panel shows the distributions in projected flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% the range of angles corresponding to
an opening angle around the
projected radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances are
needed to contrain the flight direction in three dimensions.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/Figure2_paper_vfts682_hst}  
  \caption{The thick cyan arrows indicate proper motions relative R136 (blue
    circle) from \emph{Gaia} DR2, multiplied by 1\,Myr. The prolongations in the opposite
    direction are proportional to the apparent ages of the stars, and
    the thin cyan lines illustrate the error cone. The yellow arrow
    indicates the HST proper motion of VFTS682, with the relative
    error cone and prolongation proportional to its apparent age.
    }

>>>>

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to reside in the innermost cores. In this picture, field stars are primarily the result of the dissolution of dense groups.  However, a significant population of massive stars exists in relative isolation, far from dense clusters or OB associations and their origin remains matter of debate \citep{gvaramadze:12, lamb:16,ward:18}.   One hypothesis to explain the population of relatively isolated massive stars is that they formed in the field. The alternative hypothesis is that these massive stars were ejected from the clusters in which they formed.  Such ejections may result from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems at the death of the companion  star \citep[e.g.,][]{zwicky:57, blaauw:61, eldridge:11, renzo:18}.

@@

@@

VFTS682 also appears to be relatively isolated in (near) infrared. There nearest bright (near-)infrared sources detected by \emph{Spitzer} \citep{meixner:06} and resolved in the \emph{VISTA} Magellanic Clouds Survey \citep{cioni:11} are located at a distance of about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate that these nearby young stars may represent a case of starformation triggered by the the wind of VFTS682, but we chance alignment cannot be excluded.

@@

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. The peak at $\theta\simeq\pi/2$ is due to stars
    belonging to NGC2060. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. }
  \label{fig:dist}
\end{figure*}

VFTS682 is identified with the source id 4657685637907503744 in the
\emph{Gaia} DR2 catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}
  as a 15.65 mag star in the G band
\citep{gaia:16,brown:18}.   The number of visibility periods,
i.e., groups of observations separated from other groups by a gap of at
least four days, used in the astrometric solution is seventeen for this
star. The reported astrometric excess noise is zero.  These values
suggest that the \emph{Gaia} DR2  data for VFTS682 are
reliable.

@@

\subsection{HST astrometry for VFTS682}

@@

@@

Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light shaded
blue). The green vertical bands show the expectations for VFTS682 if
it were ejected from R136. We highlight the star together with the two O-type
runaways \mbox{(re-)identified} by \cite{lennon:18} (VFTS16 and VFTS72,
respectively).

The left panel shows the distribution in two-dimensional proper motion relative
to R136. Although VFTS682 is not peculiar, its relative proper motion
falls in the range expected for it to reach its present day isolation
within its apparent age (green vertical band), hinting at it being a ``slow-runaway'' as
suggested by \cite{bestenlehner:11}. Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other, but also with no relative motion
within $2\sigma$.

% The left panel shows the distribution in two-dimensional proper motion relative
% to R136, and shows that, although VFTS682 is not a peculiar star in terms of
% its motion relative to R136. Subtracting the mean motion of the field,
% we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
% $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
% are consistent with each other and indicate
% a non-zero peculiar projected motion that would make it a slow runaway
% star as expected by \cite{bestenlehner:11} if the star was ejected
% from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
% More importantly, both the proper motion
% measurement from \emph{Gaia} and HST fall in the range expected to
% reach its present day location % at $\sim$29\,pc from the cluster core
% if the
% star was ejected very early in its evolution. This range is
% highlighted by the green vertical band, and its width is determined by
% the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

The right panel shows the distributions in projected flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% the range of angles corresponding to
an opening angle around the
projected radial direction of 45\,degrees. The direction of motion of VFTS682 % is
% poorly constrained because of the large errors on the proper motion
% components both in HST and \emph{Gaia} data, but
% it
is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

@@

\begin{figure}%[tbp]
  \centering
  \includegraphics[width=0.48\textwidth]{./figures/Figure2_paper_vfts682_hst}  
  \caption{The thick cyan arrows indicate proper motions relative R136 (blue
    circle) from \emph{Gaia} DR2, multiplied by 1\,Myr. The prolongations in the opposite
    direction are proportional to the apparent ages of the stars, and
    the thin cyan lines illustrate the error cone. The yellow arrow
    indicates the HST proper motion of VFTS682, with the relative
    error cone and prolongation proportional to its apparent
    age.}
